---
title: "Impact of severe weather on the power grid"
author: "Lauren Puffer"
format: html
editor: visual
embed-resources: true
page-loading: false
code-fold: show
execute: 
  warning: false
  message: false
toc: TRUE
theme: journal
---

# Introduction

# Data Sources

Open Street Map: <https://planet.openstreetmap.org/>

# Load packages

```{r}
#| message: false
#| warning: false 
library(terra)
library(tmap)
library(here)
library(janitor)
library(tidyverse)
library(kableExtra) # table formatting
library(spData) # spatial data
library(spDataLarge) # spatial data
library(geodata)
library(stars)
```

# Load data

## Light raster of Houston, TX

Read in raster data for Houston, TX.

```{r}
# read tile at 10:38 from Feb 7th 2021 in Houston
houston05_02_07 <- read_stars(here("data", "VNP46A1", "VNP46A1.A2021038.h08v05.001.2021039064328.tif"))

# read tile at 10:38 from Feb 7th 2021 in Houston
houston06_02_07 <- read_stars(here("data", "VNP46A1", "VNP46A1.A2021038.h08v06.001.2021039064329.tif"))

# read tile at 10:47 from Feb 16th 2021 in Houston
houston05_02_16 <- read_stars(here("data", "VNP46A1", "VNP46A1.A2021047.h08v05.001.2021048091106.tif"))

# read tile at 10:47 from Feb 16th 2021 in Houston
houston06_02_16 <- read_stars(here("data", "VNP46A1", "VNP46A1.A2021047.h08v06.001.2021048091105.tif"))
```

## Roads vector of Houston, TX

Read in vector data of roads in Houston, TX.

```{r}
# get roads as a vector
roads <- st_read(here("data", "gis_osm_roads_free_1.gpkg"), 
                 query = "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'")

# make a vector for roads
roads_vec <- vect(roads)
```

## Buildings

```{r}
# get buildings as a vector
buildings <- st_read(here("data", "gis_osm_buildings_a_free_1.gpkg"),
                     query = "SELECT * FROM gis_osm_buildings_a_free_1 WHERE (type IS NULL AND name IS NULL) OR type in ('residential', 'apartments', 'house','static_caravan', 'detached')", 
                     quiet= TRUE)

# make a vector of buildings
buildings_vec <- vect(buildings)
```

## Census data from Texas

We will need to use a datasbase that has geometries and combine it with our 'census' dataframe. Our 'census' datafram contains GEOID's which will allow us to join with another dataset conatining the correct geometries needed to check the CRS and complete our analysis.

```{r}
# get list of layers in .gdb
st_layers(here("data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"))

# load census geopackage for 2019
census_geo <- st_read(here("data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), 
                  layer = "ACS_2019_5YR_TRACT_48_Texas", 
                  quiet = TRUE)|>
  st_make_valid()

# load census income for 2019
census_income <- st_read(here("data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), 
                  layer = "X19_INCOME", 
                  quiet = TRUE)

# left join income and geometry
census_join <- left_join(census_geo, census_income, by = "GEOID")
```

# Coordinate Reference Systems

In order to ensure that all of our data is compatible spatially, we must make sure that they are all using the same Coordinate Reference System. Some spatial data is stored based on different maps or different units. For example, one dataset may be using meters for spatial data, while another is using degrees.

To ensure that our spatial data matches in CRS, we will first have to check what the CRS's are for each.

## Checking CRSs

Assuming all of the tile data are in the same CRS, we will test the CRS of one of them.

```{r}
# get CRS of Houston light data (06 tile)
st_crs(houston06_02_07)

# get CRS of Houston light data (05 tile)
st_crs(houston05_02_16)

# check that they match
st_crs(houston06_02_07) == st_crs(houston05_02_16) # they do!
```

The Houston light data uses WGS 84 (EPSG 4326). We will match all of our other data to this CRS using the st_transform() function.

Now we will transform our roads vector to match the CRS of the Houston light data (EPSG 4326).

```{r}
# check CRS of roads
st_crs(roads_vec) # roads also use EPSG 4326!

# double check that they are an exact match
st_crs(roads_vec) == st_crs(houston06_02_16) # true!
```

The roads from Open Street Map share the same CRS as the Houston light data. Now, we can check to see that the demographic data from the U.S. Census Bureau matches as well. It is likely that because the U.S. Census data and

## Change CRSs to match

```{r}
# check to see if the census data has a CRS
st_crs(census_join) == st_crs(roads_vec) # false

# check CRS of census data
st_crs(census_join)

# use st_transform() to change the CRS of census data
census_join_reproj <- st_transform (census_join, st_crs(houston05_02_07))

# check that CRSs match
st_crs(census_join_reproj) == st_crs(houston05_02_07) # true!
```

Now that all of the CRSs match for all of our data, we can begin our data manipulations to prepare them for analysis.

# Houston blackout map

To better understand which areas were impacted by the February 16th blackout, we will subtract the cell values in our light intensity tiles from before and after the blackout to get the difference.

## Combine tiles by date

We must first combine the February 7th and February 16th rasters, as they both represent different parts of the city.

```{r}
# combine feb 7th rasters
feb_7_combined <- st_mosaic(houston06_02_07, houston05_02_07)
plot(feb_7_combined)

# combine feb 16th rasters
feb_16_combined <- st_mosaic(houston06_02_16, houston05_02_16)
plot(feb_16_combined)
```

## Plot before and after

We will plot the light intensity map before and after the blackout. To do this, we will use tmap.

```{r}
# load 'viridis' package for visuals
library(viridisLite)


# plot combined Feb 7th raster
feb_7_plot <- plot(feb_7_combined,
     col = viridis(256), # color palette
     main = "Before - Feb 7, 2021",
     axes = TRUE)

# plot combined Feb 16th raster
feb_16_plot <- plot(feb_16_combined,
     col = viridis(256),
     main = "After - Feb 16, 2021",
     axes = TRUE)

# bring in patchwork to combine plots
library(patchwork)
feb_7_plot + feb_16_plot
```

## Difference in light intensity

In order to see the severity of the blackout by looking at the difference between light intensity before and after the blackout. We will do this by subtracting the cell values from the blackout date, from the cell values the week before (February 7th).

```{r}
# get the difference
difference <- feb_7_combined - feb_16_combined

# make any values with <200 an NA
difference[difference <= 200] <- NA
```

Now we need to make a vector of the difference so that we can create a mask of areas which were impacted by the blackout.

```{r}
# create a vector of the difference
mask_sf <- st_as_sf(difference, as_points = FALSE, merge = TRUE) |>
  st_make_valid()
```

Now, we will create a bounding box for the city of Houston.

```{r}
# create bbox for the city of Houston
houston_boundary <- st_bbox(c(
  xmin = -96.5,
  ymin = 29,
  xmax = -94.5,
  ymax = 30.5
), crs = st_crs(4326))  # WGS84 lon/lat

# vectorize the boundary
houston <- st_as_sfc(houston_boundary)

st_crs(houston)
```

## Crop blackout mask

We will use st_intersects() with the Houston boundary box and the difference mask to get the areas that are within the boundary.

```{r}
# get crs of mask
st_crs(mask_sf) # 4326

# use st_intersects() to get just the area of blackout difference in the bbox
blackout_houston <- mask_sf[mask_sf %>% st_intersects(houston, sparse = FALSE), ]

# show all areas where >200 difference
tm_shape(blackout_houston)+
  tm_polygons()
```

Now that we have a map of more intense blackouts, we can bring in the roads data and exclude any polygons within 200m of the road.

## Change CRS to meters

We have a roads vector which shares the same CRS as our other data, but because we want to create a buffer that is in meters, we have to change the CRS. Houston, TX uses EPSG 3083. We should be able to preserve the spatial data we have and convert to EPSG 3083 from WGS 84 (EPSG 4326).

```{r}
# check roads CRS
st_crs(roads) # EPSG 4326

# change to a CRS that uses meters
roads_trans <- st_transform(roads, crs = 3083)

# check the CRS of "blackout_houston"
st_crs(blackout_houston) # 4326

# change CRS to match "roads_trans"
blackout_houston_trans <- st_transform(blackout_houston, crs = 3083)
```

## Creating a 200m buffer around roads

Now that our roads and blackout polygons are both both in CRSs that use meters, we must add a buffer to our roads. In order to do this, we must make them into a vector. We will do this by "expanding" the geometry of the roads to 200m.

```{r}
# create 200 m buffer around roads
roads_buffer_200m <- st_buffer(roads_trans, dist = units :: set_units(200, "m"))

# use st_union() to bombine roads and buffers
roads_union <- st_union(roads_buffer_200m)
```

## Exclude polygons touching roads

Because roads are associated with larger changes in light intensity, we want to exclude them from our maps. We will do this by using st_difference() to keep only the parts of our blackout polygons that are outside of the 200 m buffer zone.

St_disjoint() would have also worked, but would have excluded any polygons touching the roads at all.

```{r}
# check classes
class(roads_union)
class(blackout_houston_trans)

# check CRS
st_crs(blackout_houston_trans) == st_crs(roads_union) # EPSG 3083

# get list of polygons that don't intersect
disjoint_index <- st_disjoint(blackout_houston_trans, roads_union)
# anything with a 0 in this list intersects with roads

# make logic out of list
disjoint_logical <- lengths(disjoint_index) != 0

# keep polygons that don't intersect with 200 m roads buffer
no_roads_blackout <- blackout_houston_trans[disjoint_logical, ]
```

# Map of homes that lost power

Now, we want to see which homes lost power from the storm. Since we already have a map of polygons with strong differences in light intensity, we can use st_intersects to keep only the homes and bulidings that overlap with the areas of strong power loss.

## Convert CRS back to EPSG 4326

Because we had to create a buffer in meters, our no_roads_blackout sf is no loner in EPSG 4326, but is instead in EPSG 3083. We will fix this by converting it back right now.

```{r}
# convert blackout map back to 4326
no_roads_blackout_trans <- st_transform(no_roads_blackout, crs = 4326)

# check to make sure it matches 'buildings'
st_crs(no_roads_blackout_trans) == st_crs(buildings) # true!
```

Now that our CRSs match, we can find homes and buildings that overlap with our blackout areas. This will allow us to get both a map of the homes that experienced blackout, and a total \# of homes that experienced blackout as well.

```{r}
# check that crs match
st_crs(no_roads_blackout_trans) == st_crs(buildings) # true

# get buildings that intersect the no_roads polygons
intersect_index <- st_intersects(buildings, no_roads_blackout_trans)

# keep buildings that have at least one intersection
blackout_buildings <- buildings[lengths(intersect_index) > 0, ]

# get the sum of homes that experienced blackout
print(nrow(blackout_buildings))
```

The total number of homes affected by the blackout was 133,804. This is also the number of observations we had in our "blackout_buildings" sf. Each building is one polygon in the 'geometry' column. 

